---
---
title: "Final Project Report"
output:
  pdf_document:
    latex_engine: xelatex

---
#Introduction

Dendritic cells (DCs) are key regulators of immune responses and are essential for initiating and shaping adaptive immunity. Among conventional dendritic cell subsets, cDC1 and cDC2 perform distinct functions: cDC1 cells are specialized for antigen cross-presentation and CD8⁺ T cell activation, whereas cDC2 cells primarily support CD4⁺ T cell–mediated responses. Proper differentiation and maintenance of these subsets depend on coordinated transcriptional and epigenetic regulation.

In the study by De Sá Fernandes et al. (2024), the authors investigated the role of histone deacetylase 1 (HDAC1) in regulating chromatin accessibility and gene expression in dendritic cells. Using an integrated multi-omics approach, including RNA-seq and ATAC-seq, they showed that loss of HDAC1 leads to widespread changes in chromatin accessibility and transcriptional programs, with a particularly strong impact on cDC2 identity. In contrast, cDC1 cells were less affected by HDAC1 deletion. In this project, we reproduced the ATAC-seq–based analyses underlying Figure 6 of the original paper to examine differential chromatin accessibility, genomic annotation, and transcription factor motif enrichment in cDC1 and cDC2 populations.


#Methods

### Quality Control

Raw sequencing reads were assessed using FastQC (v0.12.1) with default parameters. Adapter sequences and low-quality bases were removed using Trimmomatic (v0.39) with default trimming settings. Quality control outputs from FastQC, Trimmomatic, and alignment statistics were summarized using MultiQC (v1.25).

### Alignment and Processing

Trimmed reads were aligned to the mm10 mouse reference genome using Bowtie2 (v2.5.4) with ATAC-seq–appropriate alignment parameters. Reads mapping to mitochondrial DNA were removed to reduce background signal. BAM files were subsequently sorted using samtools sort and indexed using samtools index from samtools (v1.21). Mapping metrics, including alignment rates and read distributions, were collected for quality assessment.

### Coverage Tracks

Genome-wide chromatin accessibility signals were generated by converting sorted BAM files to bigWig format using a BAM-to-bigWig conversion module. These coverage tracks were used for downstream visualization, including heatmaps and aggregate signal profiles.

### Peak Calling

Chromatin accessibility peaks were called independently for each sample using MACS2 via the pipeline peak-calling module with ATAC-seq–appropriate parameters. In parallel, a BEDTools-based peak calling strategy was applied to generate BED-format peak files. These peak files were used directly for downstream differential chromatin accessibility analysis.

### Quality Control Summary

Quality metrics from FastQC, alignment statistics from Bowtie2, and post-filtering metrics following mitochondrial read removal were aggregated and summarized using MultiQC (v1.25), providing an integrated overview of data quality across all samples.




```{r}
BiocManager::install("DiffBind")
BiocManager::install("ChIPseeker")
```

```{r}
BiocManager::install("TxDb.Mmusculus.UCSC.mm10.knownGene")
BiocManager::install("org.Mm.eg.db")
```
```{r}
BiocManager::install(c(
  "chromVAR",
  "motifmatchr",
  "JASPAR2020",
  "BSgenome.Mmusculus.UCSC.mm10",
  "SummarizedExperiment"
))
```
```{r}
BiocManager::install("TFBSTools")

```


# Differential chromatin accessibility(cDC1)
Differential chromatin accessibility between WT and KO samples was analyzed separately for the cDC1 and cDC2 cell populations using the DiffBind R package. For each cell type, a DiffBind object was constructed from a sample sheet containing aligned BAM files and corresponding peak files. Read counts were summarized across a consensus peak set using a summit width of 250 bp, ensuring consistent quantification across samples.

Experimental contrasts were defined based on the condition (WT vs KO), requiring a minimum of two biological replicates per group. Differential accessibility testing was performed using DESeq2 as implemented in DiffBind. For cDC1, a relaxed reporting threshold (FDR ≤ 1) was used to capture global accessibility patterns for downstream visualization, whereas for cDC2 a more stringent threshold (FDR ≤ 0.05) was applied to identify high-confidence differentially accessible regions (DARs).

```{r}
library(DiffBind)
library(tidyverse)

```

##Load DiffBind object
```{r}
db_cDC1 <- dba(sampleSheet = "samplesheet_diffbind_cDC1.csv")
db_cDC1

```
##count reads-in-peaks
```{r}
db_cDC1 <- dba.count(
  db_cDC1,
  summits = 250
)
db_cDC1

```

##Define contrast WT vs KO
```{r}
db_cDC1 <- dba.contrast(
  db_cDC1,
  categories = DBA_CONDITION,
  minMembers = 2
)

dba.show(db_cDC1, bContrasts = TRUE)
```

##Differential analysis with DESeq2
```{r}
db_cDC1 <- dba.analyze(
  db_cDC1,
  method = DBA_DESEQ2
)
db_cDC1
```

```{r}
diff_peaks_cDC1 <- dba.report(
  db_cDC1,
  method = DBA_DESEQ2,
  th = 1
)

length(diff_peaks_cDC1)
```
#Differential chromatin accessibility — cDC2
```{r}
db_cDC2 <- dba(sampleSheet = "samplesheet_diffbind_cDC2.csv")

db_cDC2 <- dba.count(
  db_cDC2,
  summits = 250
)
db_cDC2

db_cDC2 <- dba.contrast(
  db_cDC2,
  categories = DBA_CONDITION,
  minMembers = 2
)

dba.show(db_cDC2, bContrasts = TRUE)

db_cDC2 <- dba.analyze(
  db_cDC2,
  method = DBA_DESEQ2
)
db_cDC2

diff_peaks_cDC2 <- dba.report(
  db_cDC2,
  method = DBA_DESEQ2,
  th = 0.05
)

length(diff_peaks_cDC2)

```


# Peak annotation with ChIPseeker
Differentially accessible peaks were annotated to nearby genomic features using the ChIPseeker package. Peaks were assigned to the nearest genes based on the TxDb.Mmusculus.UCSC.mm10.knownGene annotation, with transcription start site (TSS) regions defined as ±3 kb. Gene identifiers were further annotated using org.Mm.eg.db. The distribution of DARs across genomic features (e.g., promoters, introns, intergenic regions) was summarized and visualized to assess the genomic context of accessibility changes.

```{r}
library(ChIPseeker)
library(TxDb.Mmusculus.UCSC.mm10.knownGene)
library(org.Mm.eg.db)
```


##peak annotation cDC1
```{r}
anno_cDC1 <- annotatePeak(
  diff_peaks_cDC1,
  TxDb = TxDb.Mmusculus.UCSC.mm10.knownGene,
  tssRegion = c(-3000, 3000),
  annoDb = "org.Mm.eg.db"
)

plotAnnoBar(anno_cDC1)
```

##peak annotation cDC2
```{r}
anno_cDC2 <- annotatePeak(
  diff_peaks_cDC2,
  TxDb = TxDb.Mmusculus.UCSC.mm10.knownGene,
  tssRegion = c(-3000, 3000),
  annoDb = "org.Mm.eg.db"
)

plotAnnoBar(anno_cDC2)

```
Comparison of genomic feature distributions revealed that both cDC1 and cDC2 differential accessible regions are strongly enriched at promoter-proximal regions, particularly within ±1 kb of transcription start sites, indicating that chromatin accessibility changes in both cell types are closely linked to transcriptional regulation.

A notable difference between the two populations is observed in the distal intergenic fraction, which is relatively higher in cDC2 compared to cDC1. This suggests that chromatin accessibility changes in cDC2 may involve a greater contribution from distal regulatory elements, such as enhancers, whereas cDC1 accessibility changes appear to be more promoter-centric.

In both cell types, peaks annotated to exonic and UTR regions represent only a minor fraction, supporting the conclusion that differential accessibility is not randomly distributed across gene bodies but is concentrated at regulatory regions.


#Motif enrichment of differentially accessible peaks
## Motif enrichment analysis
```{r}
library(chromVAR)
library(motifmatchr)
library(JASPAR2020)
library(BSgenome.Mmusculus.UCSC.mm10)
library(SummarizedExperiment)
library(tidyverse)
library(TFBSTools)
```


```{r}
motifs <- getMatrixSet(
  JASPAR2020,
  opts = list(species = 10090, all_versions = FALSE)
)
```

## Motif matching — cDC1
```{r}
library(GenomeInfoDb)

diff_peaks_cDC1 <- keepStandardChromosomes(
  diff_peaks_cDC1,
  pruning.mode = "coarse"
)

```

```{r}
motif_ix_cDC1 <- matchMotifs(
  motifs,
  diff_peaks_cDC1,
  genome = BSgenome.Mmusculus.UCSC.mm10
)

motif_mat_cDC1 <- assay(motif_ix_cDC1)

motif_counts_cDC1 <- rowSums(motif_mat_cDC1)

top_motifs_cDC1 <- tibble(
  motif = names(motif_counts_cDC1),
  count = motif_counts_cDC1
) %>%
  arrange(desc(count)) %>%
  slice_head(n = 10)

```

```{r}
motif_matrix <- assay(motif_ix_cDC1)

dim(motif_matrix)
```

## Motif matching — cDC2
```{r}
library(GenomeInfoDb)

diff_peaks_cDC2 <- keepStandardChromosomes(
  diff_peaks_cDC2,
  pruning.mode = "coarse"
)

```

```{r}
motif_ix_cDC2 <- matchMotifs(
  motifs,
  diff_peaks_cDC2,
  genome = BSgenome.Mmusculus.UCSC.mm10
)

motif_mat_cDC2 <- assay(motif_ix_cDC2)

top_motifs_cDC2 <- tibble(
  motif = names(rowSums(motif_mat_cDC2)),
  count = rowSums(motif_mat_cDC2)
) %>%
  arrange(desc(count)) %>%
  slice_head(n = 10)

```

```{r}
motif_matrix <- assay(motif_ix_cDC2)

dim(motif_matrix)
```
#cDC1 gain / loss peaks
```{r}
diff_vis_cDC1 <- dba.report(db_cDC1, th = 1)

gain_cDC1 <- diff_vis_cDC1[mcols(diff_vis_cDC1)$Fold > 0]
loss_cDC1 <- diff_vis_cDC1[mcols(diff_vis_cDC1)$Fold < 0]

export(gain_cDC1, "cDC1_gain_peaks.bed")
export(loss_cDC1, "cDC1_loss_peaks.bed")
```

#cDC2 gain / loss peaks
```{r}
diff_vis_cDC2 <- dba.report(db_cDC2, th = 0.05)

gain_cDC2 <- diff_vis_cDC2[mcols(diff_vis_cDC2)$Fold > 0]
loss_cDC2 <- diff_vis_cDC2[mcols(diff_vis_cDC2)$Fold < 0]

export(gain_cDC2, "cDC2_gain_peaks.bed")
export(loss_cDC2, "cDC2_loss_peaks.bed")
```

```{r}
library(rtracklayer)

export(diff_vis_cDC1, "cDC1_DARs.bed")
export(diff_vis_cDC2, "cDC2_DARs.bed")

```


## figure6c — ATAC-only scatter (cDC1)
```{r}
library(tidyverse)

df_cDC1 <- as.data.frame(diff_vis_cDC1) %>%
  mutate(
    accessibility = ifelse(Fold > 0, "Gain", "Loss")
  )

ggplot(
  df_cDC1,
  aes(
    x = Fold,
    y = -log10(FDR),
    color = accessibility
  )
) +
  geom_point(alpha = 0.6, size = 1) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  scale_color_manual(
    values = c("Gain" = "#D55E00", "Loss" = "#0072B2")
  ) +
  theme_classic() +
  labs(
    title = "Differential chromatin accessibility in cDC1 (ATAC-seq)",
    x = "log2 Fold Change (ATAC-seq)",
    y = "-log10(FDR)",
    color = "Accessibility change"
  )
```


## figure6e — ATAC-only scatter (cDC2)
```{r}
library(tidyverse)

df_cDC2 <- as.data.frame(diff_vis_cDC2) %>%
  mutate(
    accessibility = ifelse(Fold > 0, "Gain", "Loss")
  )

ggplot(
  df_cDC2,
  aes(
    x = Fold,
    y = -log10(FDR),
    color = accessibility
  )
) +
  geom_point(alpha = 0.6, size = 1) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  scale_color_manual(
    values = c("Gain" = "#D55E00", "Loss" = "#0072B2")
  ) +
  theme_classic() +
  labs(
    title = "Differential chromatin accessibility in cDC2 (ATAC-seq)",
    x = "log2 Fold Change (ATAC-seq)",
    y = "-log10(FDR)",
    color = "Accessibility change"
  )

```
To assess differential chromatin accessibility, we visualized ATAC-seq log2 fold change versus statistical significance for cDC1 and cDC2 cells. In cDC1, differentially accessible regions showed relatively modest changes in accessibility, with gain and loss regions distributed symmetrically around zero, indicating subtle chromatin remodeling upon knockout. In contrast, cDC2 cells exhibited a pronounced skew toward loss of accessibility, with a large number of regions displaying strong negative log2 fold changes and high statistical significance.

While the original study integrated ATAC-seq and RNA-seq data to examine the relationship between chromatin accessibility and transcriptional output, RNA-seq data were not available for this reproduction. Therefore, our analysis focuses on ATAC-seq–only changes. Despite this limitation, the observed patterns—particularly the extensive loss of accessibility in cDC2 compared to cDC1—are consistent with the trends reported in the original figures.
